ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-09-24 paper/tables.Rmd

```{r}
res_path <- "results/"
files <- list.files(res_path)
dt <- lapply(files[grepl("_bmk.csv", files)], function(x) {
    fread(file.path(res_path, x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,ce:=NULL]
synth <- c("Moons", "Circles", "Linearly Separable")
tabular <- c("GMSC", "German Credit", "California Housing")
dt[,source:=ifelse(dataname %in% synth, "synthetic", ifelse(dataname %in% tabular, "tabular", "vision"))]
dt[,non_valid:=variable=="validity" & value==0.0,.(sample,dataname,generator,model,target,factual,source)]
dt[,non_valid:=any(non_valid==TRUE),.(sample,dataname,generator,model,target,factual,source)]
dt_valid <- dt[non_valid==FALSE]
```

@@

# Measures to be minimized:
min_measures <- c(
  "distance",
  "distance_from_energy", 
  "distance_from_targets",
  "set_size_penalty"
)
tab[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
tab[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
tab[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]

@@

# Measures to be minimized:
min_measures <- c(
  "distance",
  "distance_from_energy", 
  "distance_from_targets",
  "set_size_penalty"
)
tab_valid[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
tab_valid[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
tab_valid[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab_valid[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]

@@

```{r}
# Choices:
measures <- c("distance_from_energy", "distance_from_targets")
measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
chosen_source <- "real-world"
chosen_data <- c(
  "MNIST",
  "GMSC"
)
tab_i <- tab

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[source == chosen_source]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

```{r}
# Choices:
measures <- c("distance_from_energy", "distance_from_targets")
measure_names <- c("Unfaithfulness ↓","Implausibility ↓")
chosen_source <- "synthetic"
chosen_data <- c(
  "Linearly Separable",
  "Moons",
  "Circles"
)
tab_i <- tab

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[source == chosen_source]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

>>>>

```{r}
res_path <- "results/"
files <- list.files(res_path)
dt <- lapply(files[grepl("_bmk.csv", files)], function(x) {
    fread(file.path(res_path, x))
})
dt <- Reduce(function(x,y) {rbind(x,y, fill=TRUE)}, dt)
dt[,ce:=NULL]
synth <- c("Moons", "Circles", "Linearly Separable")
tabular <- c("GMSC", "German Credit", "California Housing")
dt[,source:=ifelse(dataname %in% synth, "synthetic", ifelse(dataname %in% tabular, "tabular", "vision"))]
```

```{r}
# Generator names
dt[,generator:=factor(generator)]
levels(dt$generator)[match("ECCCo",levels(dt$generator))] <- "ECCCo-L1"
levels(dt$generator)[match("ECCCo (no CP)",levels(dt$generator))] <- "ECCCo-L1 (no CP)"
levels(dt$generator)[match("ECCCo (no EBM)",levels(dt$generator))] <- "ECCCo-L1 (no EBM)"
levels(dt$generator)[match("ECCCo-Δ",levels(dt$generator))] <- "ECCCo"
levels(dt$generator)[match("ECCCo-Δ (latent)",levels(dt$generator))] <- "ECCCo+"
levels(dt$generator)[match("ECCCo-Δ (no CP)",levels(dt$generator))] <- "ECCCo (no CP)"
levels(dt$generator)[match("ECCCo-Δ (no EBM)",levels(dt$generator))] <- "ECCCo (no EBM)"
```

```{r}
# Adjust measure names
dt[source=="vision" & variable=="distance_from_targets_ssim", variable:="implausibility"]
dt[source=="vision" & variable=="distance_from_energy_ssim", variable:="unfaithfulness"]
dt[source!="vision" & variable=="distance_from_targets_l2", variable:="implausibility"]
dt[source!="vision" & variable=="distance_from_energy_l2", variable:="unfaithfulness"]
```

```{r}
dt[,non_valid:=variable=="validity" & value==0.0,.(sample,dataname,generator,model,target,factual,source)]
dt[,non_valid:=any(non_valid==TRUE),.(sample,dataname,generator,model,target,factual,source)]
dt_valid <- dt[non_valid==FALSE]
```

@@

# Measures to be minimized:
min_measures <- c(
  "distance",
  "implausibility",
  "unfaithfulness",
  "distance_from_energy", 
  "distance_from_energy_l2", 
  "distance_from_targets",
  "distance_from_targets_l2",
  "set_size_penalty"
)
tab[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
tab[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
tab[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]

@@

# Measures to be minimized:
min_measures <- c(
  "distance",
  "implausibility",
  "unfaithfulness",
  "distance_from_energy", 
  "distance_from_energy_l2", 
  "distance_from_targets",
  "distance_from_targets_l2",
  "set_size_penalty"
)
tab_valid[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
tab_valid[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
tab_valid[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab_valid[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]

@@

```{r}
# Choices:
measures <- c(
  "unfaithfulness",
  "implausibility"
)
measure_names <- c(
  "Unfaithfulness ↓",
  "Implausibility ↓"
)
# Order:
chosen_data <- c(
  "Linearly Separable",
  "GMSC",
  "MNIST",
)
chosen_model <- c(
  "MLP",
  "JEM",
  "LeNet-5"
)

tab_i <- tab

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i <- tab_i[model %in% chosen_model]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <-  "Results for datasets from different domains: sample averages +/- one standard deviation across counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-main} \\newline"
file_name <- "paper/contents/table-main.tex"
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = T) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

```{r}
# Choices:
measures <- c(
  "distance_from_energy_l2",
  "distance_from_targets_l2"
)
measure_names <- c(
  "Unfaithfulness ↓",
  "Implausibility ↓"
)
chosen_source <- "tabular"
# Order:
chosen_data <- c(
  "GMSC",
  "California Housing",
  "German Credit"
)
chosen_model <- "MLP"

tab_i <- tab

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[source == chosen_source]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i <- tab_i[model == chosen_model]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = T) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

```{r}
# Choices:
# Choices:
measures <- c(
  "distance_from_energy_l2",
  "distance_from_targets_l2"
)
measure_names <- c(
  "Unfaithfulness ↓",
  "Implausibility ↓"
)
chosen_source <- "synthetic"
chosen_data <- c(
  "Linearly Separable",
  "Moons",
  "Circles"
)
tab_i <- tab

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[source == chosen_source]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = T) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```

```{r}
# Choices:
measures <- c(
  "distance_from_energy_l2",
  "distance_from_targets_l2"
)
measure_names <- c(
  "Unfaithfulness ↓",
  "Implausibility ↓"
)
chosen_source <- "vision"
# Order:
chosen_data <- c(
  "MNIST",
  "Fashion MNIST",
)
chosen_model <- "LeNet-5"

tab_i <- tab

# Logic:
tab_i <- tab_i[variable %in% measures]
tab_i <- tab_i[source == chosen_source]
tab_i <- tab_i[dataname %in% chosen_data]
tab_i <- tab_i[model == chosen_model]
tab_i[,dataname:=factor(dataname,levels=chosen_data)]
tab_i <- dcast(tab_i, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(measure_names,length(chosen_data))
)
caption <- sprintf(
  "Results for %s datasets: sample averages +/- one standard deviation across counterfactuals. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-%s} \\newline",
  chosen_source,
  chosen_source
)
file_name <- sprintf(
  "paper/contents/table-%s.tex",
  chosen_source
)
sub_header <- rep(length(measures), length(chosen_data))
names(sub_header) <- chosen_data
header <- c(
  " " = 2, sub_header
)
line_sep <- c(rep("",length(measures)-1),"\\addlinespace")
algin_cols <- c(rep('l',2),rep('c',ncol(tab_i)-2))
kbl(
  tab_i, caption = caption, 
  align = algin_cols, col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = line_sep 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = T) %>%
  add_header_above(header) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable(file_name)
```
